---
title: "figure_generation"
output: html_document
date: "2025-07-18"
---

```{r setup, include=FALSE}
# Load libraries
  libraries <- c("tidyverse",
                 "magrittr",
                 "data.table",
                 "paletteer",
                 "gggenes",
                 "patchwork",
                 "lme4",
                 "ggeffects")
    
  suppressPackageStartupMessages(
    lapply(libraries, 
           require, 
           character.only = TRUE))
```

# Load data for plotting

## Load genome annotation 

1) Load genome annotation file
2) Order variables for proper plotting
3) Create amended annotation listing only gene names for plotting

```{r}

# Load genome annotation
  file <- "./reference/SARS-CoV-2_genome_annotation_NC_045512(2).csv"
  anno_db <- fread(file)
  
# Format factor variable for correct plotting order
  anno_db$gene %<>% factor(., levels = unique(.))
  anno_db$feature %<>% factor(., levels = .)
  
# Create amended anno_db with genes only for plotting
  anno_db_gene <- anno_db[, .(start = min(start),
                              end = max(end)),
                          by = gene]

```

## Set color scheme

```{r}

# Set color palette
  palette <- "wesanderson::Moonrise2"
  options(ggplot2.discrete.colour = paletteer_d(palette))

```

# Generate figures for paper

## Plot nucleotide frequency and shannon entropy by immune status

Two scatter plots showing nucleotide frequency and shannon entropy across the entire genome, combining all timepoints. 

```{r}

# Load nucleotide frequency file 
  freq_data <- fread("filtered_data/COVID_LCH_all_freq_data.csv")

# Only include alternate allele with total depth > 400
  freq_data <- freq_data[TOTAL_DP > 400]
  
# Nucleotide frequency, each dot represents an alternate allele with ALT_FREQ > 0.03
  nonzero_freq_data <- freq_data[ALT_FREQ != 0]
  
  p1 <- scatter_plot_with_genome(nonzero_freq_data,
                                 x = "POS",
                                 y = "ALT_FREQ",
                                 color = "ALT",
                                 facet_var = "immune_status",
                                 xlab = element_blank(),
                                 ylab = "Nucleotide frequency",
                                 color_lab = "Nucleotide",
                                 anno = anno_db_gene)

  
# Shannon entropy
  shannon_data <- freq_data[, 
                            shannon_entropy := calculate_shannon(ALT_FREQ), 
                            by = c("POS",
                                   "sample_ID"
                                   )]
# Only show position with positive shannon entropy values
  nonzero_shannon_data <- shannon_data[shannon_entropy != 0]
  
# If a POS has >=2 ALT, it will show as duplicate. Remove one row 
  shannon_data <- unique(shannon_data, by = c("sample_ID", "POS"))

  p2 <- scatter_plot_with_genome(nonzero_shannon_data,
                                 x = "POS",
                                 y = "shannon_entropy",
                                 color = "immune_status",
                                 facet_var = "immune_status",
                                 xlab = "Genome position",
                                 ylab = "Shannon entropy",
                                 color_lab = "Immune status",
                                 anno = anno_db_gene)
  
# Combine plots
  plot <- p1 / p2

# Save plot
  file <- "filtered_plots/nucleotide_frequency_shannon_entropy_scatter_plots_by_immune_status_400"
  ratio <- list(width = 4, height = 3)
  scale <- 2.5
  
  # Save as .tif
  ggsave(paste0(file, ".tif"), 
         plot,
         width = ratio$width,
         height = ratio$height,
         scale = scale,
         units = "in")
  
  # Save as PDF (run in command line)
  pdf(file = paste0(file, ".pdf"),
      width = ratio$width*scale,
      height = ratio$height*scale)
  
  plot(plot)
  dev.off()

```
## Plot shannon entropy as box plot per participant 

Box plots showing shannon entropy of any non-zero position at any point in the genome. Separate boxes per time point, faceted by participant ID (MRN).

```{r}

# Generate box plots
  shannon_data_timepoint<-fread("filtered_data/COVID_LCH_shannon_data_timepoint.csv")
  shannon_data_timepoint <- shannon_data_timepoint[ALT_FREQ >= 0.03]

  plot <- ggplot(data = shannon_data_timepoint,
                 aes(x = factor(Days),
                     y = shannon_entropy,
                     color = immune_status,
                     group = Days)) +
      geom_boxplot(outlier.shape = NA) +
      ggbeeswarm::geom_quasirandom(shape = 21,
                                   width = .3, 
                                   dodge.width = 0.75,
                                   aes(fill = immune_status)) +
  scale_fill_paletteer_d(palette) +
    theme_bw() +
    labs(x = "Days since first positive SARS-CoV-2 test",
         y = "Shannon entropy",
         color = "Immune status") +
    facet_wrap(vars(MRN), scales = "free_x", nrow = 3) +
    guides(fill = "none")
      

# Save plot
  file <- "filtered_plots/shannon_entropy_boxplots_by_participant_timepoint3"
  ratio <- list(width = 9, height = 5)
  scale <- 1.25
  
  # Save as .tif
  ggsave(paste0(file, ".tif"), 
         plot,
         width = ratio$width,
         height = ratio$height,
         scale = scale,
         units = "in")
  
  # Save as PDF (run in command line)
  pdf(file = paste0(file, ".pdf"),
      width = ratio$width*scale,
      height = ratio$height*scale)
  
  plot(plot)
  dev.off()

```

## Plot whole genome Shannon entropy over time

Line plot showing whole genome Shannon entropy (summed over each position) over time. Each dot represents one sample, samples from the same participant are connected by lines. Fitted curves are estimated marginal means of model. 

```{r}

# Load whole genome shannon entropy values
  wg_df <- readRDS("filtered_models/whole_genome_shannon_entropy_new.Rds")


# Generate line plot
  plot <- ggplot(data = wg_df, 
         aes(x = Days, 
             y = total_entropy,
             group = MRN,
             color = immune_status)) +
    geom_point(shape = 21,
               size = 3) +
    geom_line() +
    theme_bw() +
    labs(color = "Immune status",
         x = "Days since first positive SARS-CoV-2 test",
         y = "Total Shannon entropy")
    
  
# Get estimated marginal means for plotting
  model <- generate_model(wg_df,
                          y = "total_entropy",
                          predictors = c("immune_status",
                                         "Days",
                                         "Ct",
                                         "RDV"),
                          random = "MRN")
  
  pred <- ggpredict(model, c("Days", "immune_status"), 
                    type = "fixed",
                    ci.lvl = 0.95)
   
   # Set color scale
   cols <- paletteer_d(palette)[1:2]
   names(cols) <- levels(wg_df$immune_status)
   
# Add predicted lines to plot
   plot <- plot +
   geom_ribbon(inherit.aes = F, 
                  data = pred, 
                  aes(x = x, ymin = pred$conf.low, ymax = pred$conf.high, fill = group),
                  alpha = 0.3,
               show.legend = F) +
      geom_line(inherit.aes = F, 
                data = pred, 
                aes(x = x, y = predicted, color = group),
                linewidth = 1.5) +
    scale_color_manual(values = cols) +
    scale_fill_manual(values = cols) 
   
# Save plot
  file <- "filtered_plots/shannon_entropy_whole_genome_line_plot_timepoint7"
  ratio <- list(width = 9, height = 5)
  scale <- 1
  
  # Save as .tif
  ggsave(paste0(file, ".tif"), 
         plot,
         width = ratio$width,
         height = ratio$height,
         scale = scale,
         units = "in")
  
  # Save as PDF (run in command line)
  pdf(file = paste0(file, ".pdf"),
      width = ratio$width*scale,
      height = ratio$height*scale)
  
  plot(plot)
  dev.off()

 
```

## Plot Shannon entropy by genome feature

Box plots showing total shannon entropy across all positions within each gene, one point per sample. 

```{r}

# Load shannon entropy by gene
  gene_df <- readRDS("filtered_data/COVID_LCH_per_gene_Shannon_entropy.Rds")

  
# Plot shannon entropy boxplots by feature
  plot <- ggplot(data = gene_df,
                 aes(x = feature,
                     y = log(total_entropy+ 1),
                     color = immune_status)) +
      geom_boxplot(outlier.shape = NA) +
      ggbeeswarm::geom_quasirandom(shape = 21,
                                   width = .2, 
                                   dodge.width = 0.75,
                                   aes(fill = immune_status)) +
    scale_fill_paletteer_d(palette) +
    theme_bw() +
    labs(y = " Total Shannon entropy",
         color = "Immune status") +
    theme(axis.title.x = element_blank()) +
    guides(fill = "none")

  
# Save plot
  file <- "filtered_plots/shannon_entropy_boxplots_by_gene"
  ratio <- list(width = 2, height = 1)
  scale <- 6
  
  # Save as .tif
  ggsave(paste0(file, ".tif"), 
         plot,
         width = ratio$width,
         height = ratio$height,
         scale = scale,
         units = "in")
  
  # Save as PDF (run in command line)
  pdf(file = paste0(file, ".pdf"),
      width = ratio$width*scale,
      height = ratio$height*scale)
  
  plot(plot)
  dev.off()
  


```

